---
title: "Working with Spatial Data"
code-annotations: hover
---

```{r design-team-note}
#| include: false

# !!! NOTE TO SSECR DESIGN TEAM !!!

# This module uses raster and vector data that *are not widely available*
# You _must_ run run the 'prep-data_spatial.R' script in the "scripts/" directory before you will be able to render this module
# Reach out to Nick Lyon if questions exist
```

## Overview

Synthesis projects often have need of spatial datasets. At its simplest, it can be helpful to have a map of the original project locations including in the synthesis dataset. In more complex instances you want to extract spatial data within a certain area of sampling locations. Regardless of 'why' you're using spatial data, it may come up during your primary or synthesis work and thus deserves consideration in this course's materials. There are _many_ modes of working with spatial data, and not all of these tools require coding literacy but for consistency with the rest of the modules **this module will focus on _scripted_ approaches to interacting with spatial data**.

## Learning Objectives

After completing this topic you will be able to: 

- <u>Define</u> the difference between the two major types of spatial data
- <u>Manipulate</u> spatial data with R
- <u>Create</u> maps using spatial data
- <u>Integrate</u> spatial data with tabular data

## Needed Packages

If you'd like to follow along with the code chunks included throughout this module, you'll need to install the following packages:

```{r install-pkgs}
#| eval: false

# Note that these lines only need to be run once per computer
## So you can skip this step if you've installed these before
install.packages("tidyverse")
install.packages("sf")
install.packages("terra")
```

## Types of Spatial Data

There are two main types of spatial data: vector and raster. Both types (and the packages they require) are described in the tabs below.

:::{.panel-tabset}
### Vector Data

Vector data are stored as polygons. Essentially vector data are a set of points and--sometimes--the lines between them that define the edges of a shape. They may store additional data that is retained in a semi-tabular format that relates to the polygon(s) but isn't directly stored in them.

Common vector data types include shape files or GeoJSONs.

```{r read-vector}
#| message: false
#| output: false

# Load needed library
library(sf)

# Read in shapefile
nc_poly <- sf::st_read(dsn = file.path("data", "nc_borders.shp")) #<1>
```
1. Note that even though we're only specifying the ".shp" file in this function you _must_ also have the associated files in that same folder. In this case that includes a ".dbf", ".prj", and ".shx", though in other contexts you may have others.

Once you have read in the shapefile, you can check its structure as you would any other data object. Note that the object has both the 'data.frame' class and the 'sf' ("simple features") class. In this case, the shapefile relates to counties in North Carolina and some associated demographic data in those counties.

```{r str-vector}
# Check structure
str(nc_poly)
```

If desired, we could make a simple R base plot-style map. In this case we'll do it based on just the county areas so that the polygons are filled with a color corresponding to how large the county is.

```{r plot-vector}
#| fig-align: center

# Make a graph
plot(nc_poly["AREA"], axes = T)
```

### Raster Data

Raster data are stored as values in pixels. The resolution (i.e., size of the pixels) may differ among rasters but in all cases the data are stored at a per-pixel level.

Common raster data types include GeoTIFFs (.tif) and NetCDF (.nc) files.

```{r read-raster}
#| message: false

# Load needed library
library(terra)

# Read in raster
nc_pixel <- terra::rast(x = file.path("data", "nc_elevation.tif"))
```

Once you've read in the raster file you can check it's structure as you would any other object but the resulting output is much less informative than for other object classes.

```{r str-raster}
# Check structure
str(nc_pixel)
```

Regardless, now that we have the raster loaded we can make a simple graph to check out what sort of data is stored in it. In this case, each pixel is 3 arcseconds on each side (~0.0002° latitude/longitude) and contains the elevation (in meters) of that pixel.

```{r plot-raster}
#| fig-align: center

# Make a graph
terra::plot(nc_pixel)
```

:::

## Coordinate Reference Systems

A fundamental problem in spatial data is how to project data collected on a nearly spherical planet onto a two-dimensional plane. This has been solved--or at least clarified--by the use of <u>C</u>oordinate <u>R</u>eference <u>S</u>ystems (a.k.a. "CRS"). All spatial data have a CRS that is explicitly identified in the data and/or the metadata because the data _are not interpretable_ without knowing which CRS is used.


The CRS defines the following information:

1. **Datum** -- model for shape of the Earth including the starting coordinate pair and angular units that together define any particular point on the planet
    - Note that there can be global datums that work for any region of the world and local datums that only work for a particular area
2. **Projection** -- math for the transformation to get from a round planet to a flat map
3. **Additional parameters** -- any other information necessary to support the projection
    - E.g., the coordinates at the center of the map

Some people use the analogy of peeling a citrus fruit and flattening the peel to describe the components of a CRS. The datum is the choice between a lemon or a grapefruit (i.e., the shape of the not-quite-spherical object) while the projection is the instructions for taking the complete peel and flattening it.

You can check and transform the CRS in any scripted language that allows the loading of spatial data though the specifics do differ between the types of spatial data we introduced earlier.

:::{.panel-tabset}
### Vector CRS

For vector data we can check the CRS with other functions from the `sf` library. It can be a little difficult to parse all of the information that returns but essentially it is most important that the CRS match that of any other spatial data with which we are working.

```{r crs-check-vector}
# Check CRS
sf::st_crs(x = nc_poly)
```

Once you know the CRS, you can transform the data to another CRS if desired. This is a relatively fast operation for vector data because we're transforming geometric data rather than potentially millions of pixels.

```{r crs-transform-vector}
# Transform CRS
nc_poly_nad83 <- sf::st_transform(x = nc_poly, crs = 3083) # <1>

# Re-check CRS
sf::st_crs(nc_poly_nad83)
```
1. In order to transform to a new CRS you'll need to identify the four-digit EPSG code for the desired CRS.

### Raster CRS

For raster data we can check the CRS with other functions from the `terra` library. It can be a little difficult to parse all of the information that returns but essentially it is most important that the CRS match that of any other spatial data with which we are working.

```{r crs-check-raster}
# Check CRS
terra::crs(nc_pixel)
```

As with vector data, if desired you can transform the data to another CRS. However, unlike vector data, transforming the CRS of raster data is _very_ computationally intense. If at all possible you should avoid re-projecting rasters. If you must re-project, consider doing so in an environment with greater computing power than a typical laptop. Also, you should export a new raster in your preferred CRS after transforming so that you reduce the likelihood that you need to re-project again later in the lifecylce of your project.

In the interests of making this website reasonably quick to re-build, the following code chunk is not actually evaluated but is the correct syntax for this operation.

```{r crs-transform-raster}
#| eval: false

# Transform CRS
nc_pixel_nad83 <- terra::project(x = nc_pixel, y = "epsg:3083")

# Re-check CRS
terra::crs(nc_pixel_nad83)
```

:::

## Making Maps





## Extracting Spatial Data





## Additional Spatial Resources

### Papers & Documents

- 

### Workshops & Courses

- The Carpentries' [Introduction to Geospatial Raster & Vector Data with R](https://datacarpentry.org/r-raster-vector-geospatial/)
- The Carpentries' [Introduction to R for Geospatial Data](https://datacarpentry.org/r-intro-geospatial/index.html)
- Arctic Data Center's [Spatial and Image Data Using GeoPandas](https://learning.nceas.ucsb.edu/2023-03-arctic/sections/geopandas.html) chapter of their Scalable Computing course
- Jason Flower's (UC Santa Barbara) [Introduction to rasters with `terra`](https://jflowernet.github.io/intro-terra-ecodatascience/)
- King, R. [Spatial Data Visualization](https://github.com/king0708/spatial-data-viz) workshop

### Websites

- NASA's <u>App</u>lication for <u>E</u>xtracting and <u>E</u>xploring <u>A</u>nalysis <u>R</u>eady <u>S</u>amples [(AppEEARS) Portal](https://appeears.earthdatacloud.nasa.gov/)
