---
title: "Data Visualization & Exploration"
code-annotations: hover
---

## Overview

Under Construction, stay tuned!

## Learning Objectives

After completing this module you will be able to: 

- <u>Identify</u> appropriate graph types for given data type/distribution
- <u>Explain</u> how visualization can be a tool for quality control
- <u>Use</u> data visualization as an exploratory tool
- <u>Define</u> fundamental `ggplot2` vocabulary
- <u>Create</u> publication-quality graphs with `ggplot2`

## Needed Packages

If you'd like to follow along with the code chunks included throughout this module, you'll need to install the following packages:

```{r install-pkgs}
#| eval: false

# Note that these lines only need to be run once per computer
## So you can skip this step if you've installed these before
install.packages("tidyverse")
install.packages("cowplot")
```

We'll go ahead and load some of these libraries as well to be able to better demonstrate these concepts.

```{r libs}
#| message: false

# Load needed libraries
library(tidyverse)
library(cowplot)
```

## Module Content



## Additional Resources

### Papers & Documents

- 

### Workshops & Courses

- NCEAS Scientific Computing team's Coding in the Tidyverse workshop [`ggplot2` module](https://nceas.github.io/scicomp-workshop-tidyverse/visualize.html)
- The Carpentries' Data Analysis and Visualization in R for Ecologists [`ggplot2` episode](https://datacarpentry.org/R-ecology-lesson/04-visualization-ggplot2.html)


### Websites

- 
