## Code Demo: Post-Harmonization Visualization

After harmonizing your data, you'll want to generate one last set of 'sanity check' plots to make sure (1) you have interpreted the metadata correctly (2) you haven't made any obvious errors in the harmonization and (3) your data are ready for analysis. Nothing is less fun than finding out your analytical results are due to an error in the underlying data.

The following is a multi-part code demonstration of three common post-harmonization uses of visualization. In addition to being useful graphs, there is also example code on how to export multiple panels of graphs into separate pages of a PDF which can be really helpful when reviewing exploratory visualizations as a group (without needing to scroll through a ton of separate graph files).

If you'd like to follow along with the code chunks included throughout this demo, you'll need to install the following packages:

```{r load-pkgs}
#| message: false

## install.packages("librarian")
librarian::shelf(tidyverse, scales, ggforce, slider)
```

The three sets of plots below encompass many of the most common data structures
we have encountered types in ecological synthesis projects. These include 
quantitative measurements collected over many sites, taxonomic data collected
over many sites, and seasonal time series data.

::: panel-tabset
### Graph _All_ Numeric Variables

It can be helpful to visualize all numeric variables in your dataset, grouped by site (or dataset source) to check that the data have been homogenized correctly. As an example, we'll use a 2019 dataset on lake water quality, chemistry, and zooplankton community composition near the [Niwot Ridge](https://nwt.lternet.edu/) LTER. The dataset is a survey of 16 high alpine lakes and has structure similar to one that might be included in a multi-site synthesis. For more information on these data, check out [the data package on EDI](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-nwt.12.1). 

```{r demo_all-num-vars_data}
# Read in data
green_biochem <- read.csv(file = file.path("data", "green-lakes_water-chem-zooplank.csv")) %>%  # <1>
  dplyr::mutate(date = as.Date(date))

# Check structure
str(green_biochem)
```
1. Note that you could also read in this data directly from EDI. See ~line 31 of [this script](https://github.com/lter/ssecr/blob/main/scripts/prep-data_data-viz-demo.R) for a syntax example

Once we have the data file, we can programmatically identify all columns that R knows to be numeric.

```{r demo_all-num-vars_numcols}
# determine which columns are numeric in green_biochem
numcols <- green_biochem %>%
  dplyr::select(dplyr::where(~ is.numeric(.x) == TRUE)) %>% # <1>
  names(.) %>% 
  sort(.)

# Check that out
numcols # <2>
```
1. The tilde (`~`) is allowing us to evaluate each column against this conditional
2. You may notice that these columns all have `"num"` next to them in their structure check. The scripted method is _dramatically_ faster and more reproducible than writing these names down by hand

Now that we have our data and a vector of numeric column names, we can generate a multi-page PDF of scatterplots where each page is specific to a numeric variable and each graph panel within a given page reflects the time series at each site.

```{r demo_all-num-vars_viz-code-fake}
#| eval: false

# Open PDF 'device'
grDevices::pdf(file = file.path("qc_all_numeric.pdf")) # <1>

# Loop across numeric variables
for (var in numcols) {
  
  # Create a set of graphs for one variable
  myplot <- ggplot(green_biochem, aes(x = date, y = .data[[var]])) +
    geom_point(alpha = 0.5) + # <2>
    facet_wrap(. ~ local_site)
  
  # Print that variable
  print(myplot)
}

# Close the device
dev.off() # <3>
```
1. This function tells R that the following code should be saved as a PDF
2. A scatterplot may not be the best tool for your data; adjust appropriately
3. This function (when used after a 'device' function like `grDevices::pdf`) tells R when to stop adding things to the PDF and actually save it

The first page of the resulting plot should look something like the following, with each page having the same content but a different variable on the Y axis.

```{r demo_all-num-vars_viz-code-real}
#| fig-align: center
#| fig-width: 8
#| fig-height: 8
#| echo: false
#| warning: false

# Grab just one variable
var <- sort(numcols)[1]

# Graph it
ggplot(green_biochem, aes(x = date, y = .data[[var]])) +
  geom_point(alpha = 0.5) +
  facet_wrap(~local_site, ncol = 3)
```

### Taxonomic Consistency

Taxonomic time series can be tricky to work with due to inconsistencies in nomenclature and/or sampling effort. In particular, 'pseudoturnover' where one species 'disappears' with or without the simultaneous 'appearance' of another taxa can be indicative of either true extinctions, or changes in species names, or changes in methodology that cause particular taxa not to be detected. A second complication is that taxonomic data are often archived as 'presence-only' so it is necessary to _infer_ the absences based on sampling methodology and add them to your dataset before analysis.

While there are doubtless many field-collected datasets that have this issue, we've elected to simulate data so that we can emphasize the visualization elements of this problem while avoiding the "noise" typical of real data. This simulation is not necessarily vital to the visualization so we've left it out of the following demo. _However_, if that is of interest to you, see [this script](https://github.com/lter/ssecr/blob/main/scripts/prep-data_data-viz-demo.R)--in particular \~line 41 through \~80.

A workflow for establishing taxonomic consistency and plotting the results is included below.

```{r demo_tax-consist_data}
# Read in data
taxa_df <- read.csv(file.path("data", "simulated-taxa-df.csv"))

# Check structure
str(taxa_df)
```

First, we'll define units of sampling (year, plot and taxon) and 'pad out' the zeros. In this example, we have only added zeroes for taxa-plot-year combinations where that taxa is present in at least one year at a given plot. Again, this zero-padding is prerequisite to the visualization but not necessarily part of it so see \~lines 84-117 of the [prep script](https://github.com/lter/ssecr/blob/main/scripts/prep-data_data-viz-demo.R) if that process is of interest.

```{r  demo_tax-consist_data-2}
# Read in data
withzeros <- read.csv(file.path("data", "simulated-taxa-df_with-zeros.csv")) %>% 
  dplyr::mutate(plot = factor(plot))

# Check structure
str(withzeros) # <1>
```
1. Notice how there are more rows than the preceding data object and several new zeroes in the first few rows?

Now that we have the data in the format we need, we'll create a plot of species counts over time with zeros filled in. Because there are many plots and it is difficult to see so many panels on the same page, we'll use the `facet_wrap_paginate` function from the `ggforce` package to create a multi-page PDF output.

```{r demo_tax-consist_viz-code-fake}
#| eval: false

# Create the plot of species counts over time (with zeros filled in)
myplot <- ggplot(withzeros, aes(x = year, y = n, group = plot, color = plot)) +
  geom_line() +
  scale_x_continuous(breaks = scales::pretty_breaks()) +
  ggforce::facet_wrap_paginate(~ taxon, nrow = 2, ncol = 2) # <1>

# Start the PDF output
grDevices::pdf(file.path("counts_by_taxon_with_zeros.pdf"),
               width = 9, height = 5)

# Loop across pages (defined by `ggforce::facet_wrap_paginate`)
for (i in seq_along(ggforce::n_pages(myplot))) {
  
  page_plot <- myplot + 
      ggforce::facet_wrap_paginate(~taxon, page = i, 
                                   nrow = 2, ncol = 2)
  
  print(page_plot)
}

# Close the PDF output
dev.off()
```
1. This allows a faceted graph to spread across more than one page. See `?ggforce::facet_wrap_paginate` for details

The first page of the resulting plot should look something like this:

```{r demo_tax-consist_viz-code-real}
#| fig-align: center
#| fig-width: 7
#| fig-height: 5
#| message: false
#| echo: false

# Make multi-page graph
myplot <- ggplot(withzeros, aes(x = year, y = n, group = plot, color = plot)) +
  geom_line() +
  scale_x_continuous(breaks = scales::pretty_breaks()) +
  ggforce::facet_wrap_paginate(~ taxon, nrow = 2, ncol = 2)

# Check out just the first page
myplot +
  ggforce::facet_wrap_paginate(~ taxon, page = 1,
                               nrow = 2, ncol = 2)
```

Notice how "Taxon_A" is absent from all plots in 2014 whereas "Taxon_B" has extremely high counts in the same year. Often this can signify inconsistent use of taxonomic names over time.

### Seasonal Time Series

For time series, intra-annual variation can often make data issues difficult to spot. In these cases, it can be helpful to plot each year onto the same figure and compare trends across study years.

As an example, we'll use a 2024 dataset on streamflow near the [Niwot Ridge](https://nwt.lternet.edu/) LTER. The dataset is a 22 year time-series of daily streamflow. For more information on these data, check out [the data package on EDI](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-nwt.105.18). 

```{r demo_seasons_data}
# Read data
green_streamflow <- read.csv(file.path("data", "green-lakes_streamflow.csv")) # <1>

# Check structure
str(green_streamflow)
```
1. Note again that you could also read in this data directly from EDI. See ~line 129 of [this script](https://github.com/lter/ssecr/blob/main/scripts/prep-data_data-viz-demo.R) for a syntax example

Let's now calculate a moving average encompassing the 5 values before and after each focal value.

```{r demo_seasons_wrangle}
#| message: false

# Do necessary wrangling
stream_data <- green_streamflow %>%
  # Calculate moving average for each numeric variable
  dplyr::mutate(dplyr::across(.cols = dplyr::all_of(c("discharge", "temperature")),
                              .fns = ~ slider::slide_dbl(.x = .x, .f = mean,
                                                         .before = 5, .after = 5),
                              .names = "{.col}_move.avg" )) %>%
  # Handle date format issues
  dplyr::mutate(yday = lubridate::yday(date),
                year = lubridate::year(date))

# Check the structure of that
str(stream_data)
```

Plot seasonal time series of each numeric variable as points with the moving
average included as lines

```{r demo_seasons_viz-code-fake}
#| eval: false

# Start PDF output
grDevices::pdf(file = file.path("qc_all_numeric_seasonal.pdf"))

# Loop across variables
for (var in c("discharge", "temperature")) {
  
  # Make the graph
  myplot <- ggplot(stream_data, aes(x = yday, group = year, color = year)) +
    geom_point(aes(y = .data[[var]])) + # <1>
    geom_line(aes(y = .data[[paste0(var, "_move.avg")]])) + # <2>
    viridis::scale_color_viridis()
  
  # Print it
  print(myplot)
}

# End PDF creation
dev.off()
```
1. Add points based on the year
2. Adding lines based on the average

The first page of the resulting figure should look something like this:

```{r demo_seasons_viz-code-real}
#| fig-align: center
#| fig-width: 8
#| fig-height: 8
#| echo: false
#| warning: false

# Print the first graph
ggplot(stream_data, aes(x = yday, group = year, color = year)) +
  geom_point(aes(y = .data[["discharge"]])) +
  geom_line(aes(y = .data[["discharge_move.avg"]])) +
  viridis::scale_color_viridis()
```

One of these years is not like the others...
:::
