## Conditionals

:::{.callout-tip icon="false" collapse="false"}
### {{< fa brain >}} Learning Objectives

After completing this topic you will be able to: 

- <u>Describe</u> what a "conditional" is and how it's used in code
- <u>Use</u> `dplyr::case_when()` to simplify conditionals with many possible options

:::

Rather than finding and replacing content, you may want to create a new column based on the contents of a different column. In plain language you might phrase this as 'if column X has \[some values\] then column Y should have \[other values\]'. These operations are called <u>conditionals</u> and are an important part of data wrangling.

If you only want your conditional to support two outcomes (as in an either/or statement) there are useful functions that support this. Let's return to our Plum Island Ecosystems crab dataset for an example.

```{r conditional-libs}
#| include: false

# Load the lterdatasampler package
library(lterdatasampler)

# Load the fiddler crab dataset
data(pie_crab)
```

```{r ifelse}
#| message: false

# Load tidyverse
library(tidyverse)

# Make a new colum with an either/or conditional
pie_crab_v2 <- pie_crab %>% 
  dplyr::mutate(size_category = ifelse(test = (size >= 15), # <1>
                                       yes = "big",
                                       no = "small"),
                .after = size) 

# Count the number of crabs in each category
pie_crab_v2 %>% 
  dplyr::group_by(size_category) %>% 
  dplyr::summarize(crab_ct = dplyr::n())
```

1.  `mutate` makes a new column, `ifelse` is actually doing the conditional

If you have multiple different conditions you _can_ just stack these either/or conditionals together but this gets cumbersome quickly. It is preferable to instead use a function that supports as many alternates as you want!

```{r case-when}
# Make a new column with several conditionals
pie_crab_v2 <- pie_crab %>% 
  dplyr::mutate(size_category = dplyr::case_when( 
    size <= 10 ~ "tiny", # <1>
    size > 10 & size <= 15 ~ "small",
    size > 15 & size <= 20 ~ "big",
    size > 20 ~ "huge",
    TRUE ~ "uncategorized"), # <2>
                .after = size)

# Count the number of crabs in each category
pie_crab_v2 %>% 
  dplyr::group_by(size_category) %>% 
  dplyr::summarize(crab_ct = dplyr::n())
```

1.  Syntax is 'test \~ what to do when true'
2.  This line is a catch-all for any rows that _don't_ meet previous conditions

Note that you can use functions like this one when you do have an either/or conditional if you prefer this format.
