## Loops

:::{.callout-tip icon="false" collapse="false"}
### {{< fa brain >}} Learning Objectives

After completing this topic you will be able to:

- <u>Identify</u> the core anatomy of a "for loop"

:::

Another way of simplifying repetitive operations is to use a "for loop" (often called simply "loops"). Loops allow you to iterate across a piece of code for a set number of times. Loops require you to define an "index" object that will change itself at the end of each iteration of the loop before beginning the next iteration. This index object's identity will be determined by whatever set of values you define at the top of the loop.

Here's a very bare bones example to demonstrate the fundamentals.

```{r loop-1}
# Loop across each number between 2 and 4
for(k in 2:4){ # <1>
  
  # Square the number
  result <- k^2
  
  # Message that outside of the loop
  message(k, " squared is ", result)
} # <2>
```

1.  'k' is our index object in this loop
2.  Note that the operations to iterate across are wrapped in curly braces (`{...}`)

Once you get the hang of loops, they can be a nice way of simplifying your code in a relatively human-readable way! Let's return to our Plum Island Ecosystems crab dataset for a more nuanced example.

```{r loop-2}
#| eval: false

# Create an empty list
crab_list <- list()

# Let's loop across size categories of crab
for(focal_size in unique(pie_crab_v4$size_category)){ # <1>
  
  # Subset the data to just this size category
  focal_df <- pie_crab_v4 %>% 
    dplyr::filter(size_category == focal_size)
  
  # Calculate average and standard deviation of size within this category
  size_avg <- mean(focal_df$size, na.rm = T) 
  size_dev <- sd(focal_df$size, na.rm = T) 
  
  # Assemble this into a data table and add to our list
  crab_list[[focal_size]] <- data.frame("size_category" = focal_size,
                                        "size_mean" = size_avg,
                                        "size_sd" = size_dev)
} # Close loop

# Unlist the outputs into a dataframe
crab_df <- purrr::list_rbind(x = crab_list) # <2>

# Check out the resulting data table
crab_df
```

1.  Note that this is not the most efficient way of doing group-wise summarization but is--hopefully--a nice demonstration of loops!
2.  When all elements of your list have the same column names, `list_rbind` efficiently stacks those elements into one longer data table.
