## Custom Functions

Writing your own, customized functions can be really useful particularly when doing synthesis work. Custom functions are generally useful for reducing duplication and increasing ease of maintenance and also can be useful end products of synthesis work in and of themselves.

If one of your group's outputs is a new standard data format or analytical workflow, the functions that you develop to aid yourself become valuable to anyone who adopts your synthesis project's findings into their own workflows. If you get enough functions you can even release a package that others can install and use on their own computers. Such packages are a valuable product of synthesis efforts and can be a nice addition to a robust scientific resume/CV.

```{r custom-fxns}
#| fig-align: center
#| fig-width: 4
#| fig-height: 3

# Define custom function
crab_hist <- function(df, size_cat){
  
  # Subset data to the desired category
  data_sub <- dplyr::filter(.data = df, size_category == size_cat)
  
  # Create a histogram
  p <- psych::multi.hist(x = data_sub$size)
}

# Invoke function
crab_hist(df = pie_crab_v4, size_cat = "tiny")
```

When writing your own functions it can also be useful to program defensively. This involves anticipating likely errors and writing your own error messages that are more informative to the user than whatever machine-generated error would otherwise get generated

```{r custom-fxns-improved}
#| fig-align: center
#| fig-width: 4
#| fig-height: 3

# Define custom function
crab_hist <- function(df, size_cat = "small"){ # <1>
  
  # Error out if 'df' isn't the right format
  if(is.data.frame(df) != TRUE) # <2>
    stop("'df' must be provided as a data frame")
  
  # Error out if the data doesn't have the right columns
  if(all(c("size_category", "size") %in% names(df)) != TRUE) # <3>
    stop("'df' must include a 'size' and 'size_category' column")
  
  # Error out for unsupported size category values
  if(size_cat %in% unique(df$size_category) != TRUE)
    stop("Specified 'size_cat' not found in provided data")
  
  # Subset data to the desired category
  data_sub <- dplyr::filter(.data = df, size_category == size_cat)
  
  # Create a histogram
  p <- psych::multi.hist(x = data_sub$size)
}

# Invoke new-and-improved function
crab_hist(df = pie_crab_v4) # <4>
```

1.  The default category is now set to "small"
2.  We recommend phrasing your error checks with this format (i.e., 'if \<some condition\> is *not* true, then \<informative error/warning message\>)
3.  The `%in%` operator lets you check whether one value matches any element of a set of accepted values. Very useful in contexts like this because the alternative would be a lot of separate "or" conditionals
4.  We don't need to specify the 'size_cat' argument because we can rely on the default
