## Custom Functions
 
:::{.callout-tip icon="false" collapse="false"}
### {{< fa brain >}} Learning Objectives

After completing this topic you will be able to: 

- <u>Describe</u> the process of writing a new function
- <u>Identify</u> the core syntax required for writing a function in R
- <u>Write</u> your own function
- <u>Define</u> "defensive" coding in the context of functions
- <u>Create</u> custom warning and error messages for predictable error states

:::

Writing your own functions can be really useful, particularly when doing synthesis work. Custom functions are generally useful for reducing duplication and increasing ease of maintenance. They may also become end products of synthesis work in and of themselves!

If one of your group's outputs is a new standard data format or analytical workflow, the functions that you develop to aid yourself become valuable to anyone who adopts your synthesis project's findings into their own workflows. If you get enough functions you can even release a package that others can install and use on their own computers. Such packages are a valuable product of synthesis efforts and can be a nice addition to a robust scientific resume or CV.

### Core Function Writing Process

There are a number of ways one might go about writing a function, this topic will focus on one that is fairly flexible and should be applicable to most contexts you're likely to run into in a synthesis context.

To make this more tangible, let's load a dataset from the `lterdatasampler` package. We'll use a dataset collected at the Niwot Ridge LTER on American Pika (_Ochotona princeps_) stress and habitat. You can see the full metadata for this dataset on the Environmental Data Initiative's (EDI) website (Package ID: [knb-lter-nwt.268.1](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-nwt.268.1)).

```{r fxn-tutorial_load-data}
#| message: false

# Load the needed libraries (install if needed)
## install.packages("librarian")
librarian::shelf(lterdatasampler, dplyr)

# Load the dataset and assign to an object
pika_df <- lterdatasampler::nwt_pikas # <1>

# Check the structure
str(pika_df)
```
1. Just doing this to have a shorter object name to refer to as we continue

:::{.panel-tabset}
#### 1. Write Normal Code

The first step of writing your own function is something you'd likely already do: **write normal, non-function code that does what you want!**

In this case, let's say we're interested in extracting the results of a simple linear model into a tidy table. This is a classic context for a custom function because we'll fit _many_ models and likely will want to just copy/paste those results into our manuscript.

```{r fxn-tutorial_step-01}
# Fit one linear model
pika_mod <- lm(concentration_pg_g ~ site, data = pika_df) # <1>

# Get the summary of the model and assign that to an object
pika_result <- summary(pika_mod)

# Extract the coefficient table as a dataframe
pika_coef <- as.data.frame(pika_result$coefficients)

# The terms are preserved as row names but we'll need them as columns
pika_coef2 <- pika_coef |> # <2>
  dplyr::mutate(terms = rownames(pika_coef),
    .before = dplyr::everything()) |> # <3>
  # Rename columns to avoid spaces and special characters
  dplyr::rename(estimate = Estimate,
    std.error = `Std. Error`,
    t.value = `t value`,
    p.value = `Pr(>|t|)`) |>
  # Grab theadjusted R2 as well
  dplyr::mutate(adj.r.squared = pika_result$adj.r.squared) # <4>

# Finally, ditch the superseded row names
rownames(pika_coef2) <- NULL

# Does that look like what we want?
pika_coef2
```
1. Here, we are testing whether the concentration of a stress-related metabolite differs among sites
2. We're using R's "native pipe" (`|>`) to avoid `magrittr` as a "dependency" (i.e., a package required by our function)
3. **<u>Namespace _ALL_ functions</u>** when you're planning on writing a function so that there's no chance that the order in which a user loads their packages breaks your function. It's okay to skip base R function namespacing for brevity's sake
4. Note the adjusted R^2^ is found in a different part of the _result_ object (rather than the _coefficient_ object where the rest of the results table can be found)

<details>
<summary>Step 1: Nitty Gritty</summary>

In order to write something like the above, **you'll (likely) need to do some experimenting to figure out _where_ in each data object the information you want lives!** This might look something like the following.

```{r fxn-tutorial_step-01_sausage}
# Fit one linear model
pika_mod <- lm(concentration_pg_g ~ site, data = pika_df)

# Get the summary of the model and assign that to an object
pika_result <- summary(pika_mod)

# Check the structure of the 'results' object
str(pika_result) # <1>
```
1. Here we can see the "coefficients" element has what we want but in some cases you may need to manually check each element (i.e., by running `str`) until you find the one(s) that have what you want

</details>

#### 2. Identify Arguments

Once we've written the code that does what we want, **the next step is to identify what pieces of the code the user of our function _should have control over_.** These pieces will become the "arguments" (a.k.a. "parameters") in our function.

For our function code, the user should be able to:

1. **Supply any linear model returned by `lm`**
    - Accepting the object returned by `lm` means we don't need to have an argument in our function for every possible argument in `lm`. We can just accept a single object and work with that--that'll be much easier for us!
2. **Decide whether or not the adjusted R^2^ is returned**
    - This is not necessarily critical to the intent of our function but by giving the user control here, we can demonstrate some other nice features of custom functions

Note that this step may not require writing any code! You can just look at your code from step 1 and think critically about what 'levers' you want to give the user.


#### 3. Translate Into a Function

Once we've written normal code and identified what our arguments should be, **we need to translate our code into R's function syntax.**

Function syntax requires a few critical pieces; each will be flagged with a number footnote on the right side of the code chunk. <u>Hover over those numbers for more details</u>.

Note that it's good practice to give your function a name that is a verb so the user can get as much information as possible from just the function name.

```{r fxn-tutorial_step-03-define}
# Define the function
result_extract <- function( # <1>
  model = NULL, # <2>
  include_r2 = TRUE){ 

  # Extract the summary
  result_list <- summary(model) # <3>

  # Extract the coefficient table
  coef_df <- as.data.frame(result_list$coefficients)

  # Process that table as we did before
  ## We'll exclude comments for brevity but you totally could include comments!
  coef.tidy_df <- coef_df |>
    dplyr::mutate(terms = rownames(coef_df),
      .before = dplyr::everything()) |>
    dplyr::rename(estimate = Estimate,
      std.error = `Std. Error`,
      t.value = `t value`,
      p.value = `Pr(>|t|)`)

  # Add R2 if requested
  if(include_r2 == TRUE){
    result <- dplyr::mutate(.data = coef.tidy_df, 
      adj.r.squared = result_list$adj.r.squared)

  # If not requested, just duplicate the tidy coefficient table with a new name
  } else { result <- coef.tidy_df } # <4>

  # Remove rownames
  rownames(result) <- NULL

  # Return that to the user
  return(result) # <5>
} # <6>
```
1. We use the `function` function to create a new function
2. Inside the `function` parentheses (`(`/`)`), we define what arguments the function uses and can set default values! `NULL` is a great default so that the function will obviously break if the user happens to have an object of the same name in their script already
3. As we translate our step 1 code into the function, **<u>be sure to replace your original object names with their respective argument names!</u>**
4. Critical that the final object has the same name, _regardless_ of which side of the conditional is activated
5. By default, an R function will return the last object invoked. Relying on that default is _risky!_ It is good practice to **use the `return` function to be explicit about what the function gives back to the user.**
6. End your function with a closing curly brace (`}`). This syntax may be familiar to you if you've used `for` loops or written conditionals before.

```{r fxn-tutorial_step-03-invoke}
# Fit a linear model
pika_mod <- lm(concentration_pg_g ~ site, data = pika_df)

# Invoke our new custom function!
result_extract(model = pika_mod, include_r2 = FALSE)
```

Looks like that worked! Does it still work with different terms supplied to the linear model?

```{r fxn-tutorial_step-03-invoke2}
# Let's test elevation directly instead of site
pika_mod2 <- lm(concentration_pg_g ~ elev_m, data = pika_df)

# Invoke our new custom function!
result_extract(model = pika_mod2) # <7>
```
7. We can exclude the `include_r2` argument if we want the default behavior to be used

:::

### Coding Defensively

If you follow the above steps, you should have a working function so you _could_ go about your business and leave it at that. _However_, **it is good practice to add your own custom error and warning messages.** That way, if your function breaks, the user has some useful hints on how they can fix it on their end without being mad at you for writing a bad function.

Writing code in this way is known as programming "defensively." This involves anticipating likely errors and writing your own error messages that are more informative to the user than whatever machine-generated error would otherwise get generated.

Just like the process of writing the function in the first place, you can follow some steps to make sure your functions are as robust as possible.

:::{.panel-tabset}
#### 4. Consider Likely Failure Points

The first step of programming defensively (fourth overall in writing functions), is **to think about what errors a reasonable user is likely to encounter.** You don't need to anticipate _everything,_ just think through some of the more 'obvious' possible breaking points.

A good tip for this is to **consider how a user could break your arguments.** That is how they'll interact with your function so it is also their means of accidentally breaking it.

In our function from earlier, there are two likely breaking points:

1. **They supply a model returned from some statistics function other than `lm`**
    - Our function relies _entirely_ on the _exact_ internal structure of the object returned by `lm`. The odds that a different function returns the same internal components (so that our function would still work) are basically zero so we need to account for that process
2. **They supply a weird value to the `include_r2` argument**
    - Because we wrote the function _we_ know that the `include_r2` argument needs to be either `TRUE` or `FALSE` to be interpreted correctly. However, some other user might think they can supply some other value (e.g., "yes", "not-adjusted") and our code would not know what to do with that.

#### 5. Add Checks

Once we have thought through the potential failure points, we can **add checks inside the function to catch those and return <u>_informative_</u> messages to the user.**

Keep in mind, we _know_ what the function requires! That means we can write our checks to say 'if \<some condition\> is _NOT_ true, then \<informative error/warning message\'. In R, that means we'll use the `stop` and `warning` functions (for errors and warnings, respectively).

```{r fxn-tutorial_step-05-add-checks}
# Define the function
result_extract <- function(model = NULL, include_r2 = TRUE){ 

  # Error checks for 'model' argument
  if(is.null(model) || class(model) != "lm") # <1>
    stop("'model' must be an object returned by `stats::lm`")

  # Warning check for 'include_r2' argument
  if(is.logical(include_r2) != TRUE || length(include_r2) != 1){ # <2>
    warning("'include_r2' must be a single logical. Coercing to TRUE")
    include_r2 <- TRUE 
  } # <3>

  # Extract the summary
  result_list <- summary(model)

  # Extract the coefficient table
  coef_df <- as.data.frame(result_list$coefficients)

  # Process that table as we did before
  coef.tidy_df <- coef_df |>
    dplyr::mutate(terms = rownames(coef_df),
      .before = dplyr::everything()) |>
    dplyr::rename(estimate = Estimate,
      std.error = `Std. Error`,
      t.value = `t value`,
      p.value = `Pr(>|t|)`)

  # Add R2 if requested
  if(include_r2 == TRUE){
    result <- dplyr::mutate(.data = coef.tidy_df, 
      adj.r.squared = result_list$adj.r.squared)

  # If not requested, just duplicate the tidy coefficient table with a new name
  } else { result <- coef.tidy_df }

  # Remove rownames
  rownames(result) <- NULL

  # Return that to the user
  return(result)
} 
```
1. The `||` operator is a special conditional that returns `TRUE` <u>as soon as the _first_ condition returns `TRUE` **and** doesn't try to compute any other conditions in the same `if`</u>. This is super useful because if the `model` argument _is_ `NULL`, then we already know the `class` check will fail so we don't need to even test it. You can use this property to list out your conditions for failure from most general to most precise to save computing time on arguments that receive larger data objects or where computing the check itself is computationally-intensive. **You want the function to fail early if it is going to fail anyway!**
2. For arguments that we know have to be either `TRUE` or `FALSE`, we can skip some of the other checks and jump straight to asking "is this a logical value?" and then return a warning/error if it isn't.
3. Here we use curly braces (`{`/`}`) after the `if` because we want to _both_ return an informative message _and_ re-set the argument to an appropriate value. R doesn't do "informative indents" like some other languages (e.g., Python), so if we want multiple operations to trigger from the same condition, we need to wrap them in curly braces.

Once we've added some custom error checks, if we re-run the function with entries we _know_ are bad, we can see that our custom error/warning messages show up!

```{r fxn-tutorial_step-05-invoke}
#| eval: false
# Invoke our function on the wrong value passed to 'model'
result_extract(model = "banana", include_r2 = "please")
```

```
Error in `result_extract()`:
! 'model' must be an object returned by `stats::lm`
```

```{r fxn-tutorial_step-05-invoke2}
# Invoke our function on the elevation model with a bad value passed to 'include_r2'
result_extract(model = pika_mod2, include_r2 = c(TRUE, FALSE))
```

:::

