## Uniting or Separating Columns

Sometimes one column has multiple pieces of information that you'd like to consider separately. A date column is a common example of this because particular months are always in a given season regardless of the specific day or year. So, it can be useful to break a complete date (i.e., year/month/day) into its component bits to be better able to access those pieces of information.

```{r separate-wider-delim}
# Split date into each piece of temporal info
pie_crab_v3 <- pie_crab_v2 %>% 
  tidyr::separate_wider_delim(cols = date, 
                              delim = "-", # <1>
                              names = c("year", "month", "day"),
                              cols_remove = TRUE) # <2>

# Check that out
str(pie_crab_v3)
```

1.  'delim' is short for "delimiter" which we covered in the Reproducibility module
2.  This argument specifies whether to remove the original column when making the new columns

While breaking apart a column's contents can be useful, it can also be helpful to combine the contents of several columns!

```{r unite}
# Re-combine data information back into date
pie_crab_v4 <- pie_crab_v3 %>% 
  tidyr::unite(col = "date",
               sep = "/", # <1>
               year:day, 
               remove = FALSE) # <2>

# Structure check
str(pie_crab_v4)
```

1.  This is equivalent to the 'delim' argument in the previous function
2.  Comparable to the 'cols_remove' argument in the previous function

Note in this output how despite re-combining data information the column is listed as a character column! Simply combining or separating data is not always enough so you need to really lean into frequent data structure checks to be sure that your data are structured in the way that you want.
