## Joining Data

Often the early steps of a synthesis project involve combining the data tables horizontally. You might imagine that you have two groups' data on sea star abundance and--once you've synonymized the column names--you can simply 'stack' the tables on top of one another. Slightly trickier but no less common is combining tables by the contents of a shared column (or columns). Cases like this include wanting to combine your sea star table with ocean temperature data from the region of each group's research. You can't simply attach the columns because that assumes that the row order is identical between the two data tables (and indeed, that there are the same number of rows in both to begin with!). In this case, if both data tables shared some columns (perhaps "site" and coordinate columns) you can use **joins** to let your computer match these key columns and make sure that only appropriate rows are combined.

Because joins are completely dependent upon the value in both columns being an *exact* match, it is a good idea to carefully check the contents of those columns before attempting a join to make sure that the join will be successful.

```{r diff-check-1}
#| eval: false

# Create a fish taxonomy dataframe that corresponds with the earlier fish dataframe
fish_tax <- data.frame("species" = c("salmon", "bass", "halibut", "eel"),
                       "family" = c("Salmonidae", "Serranidae", "Pleuronectidae", "Muraenidae"))

# Check to make sure that the 'species' column matches between both tables
supportR::diff_check(old = fish_ct$species, new = fish_tax$species) 
```

```{r diff-check-2}
#| eval: false

# Use text replacement methods to fix that mistake in one table
fish_tax_v2 <- fish_tax %>% 
  dplyr::mutate(species = gsub(pattern = "^eel$", # <1>
                               replacement = "moray eel", 
                               x = species))

# Re-check to make sure that fixed it
supportR::diff_check(old = fish_ct$species, new = fish_tax_v2$species)
```

1.  The symbols around "eel" mean that we're only finding/replacing *exact* matches. It doesn't matter in this context but often replacing a partial match would result in more problems. For example, replacing "eel" with "moray eel" could make "electric eel" into "electric moray eel".

Now that the shared column matches between the two two dataframes we can use a join to combine them! There are four types of join:

1.  left/right join
2.  full join (a.k.a. outer join)
3.  inner join
4.  anti join

You can learn more about the types of join [here](https://lter.github.io/workshop-tidyverse/join.html) or [here](https://njlyon0.github.io/teach_r-for-biologists/materials/slides_4a.html#/title-slide) but the quick explanation is that <u>the name of the join indicates whether the rows of the "left" and/or the "right" table are retained in the combined table</u>. In synthesis work a left join or full join is most common (where you have your primary data in the left position and some ancillary/supplementary dataset in the right position).

```{r join}
#| eval: false

# Let's combine the fish count and fish taxonomy information
fish_df <- fish_ct %>% 
  # Actual join step
  dplyr::left_join(y = fish_tax_v2, by = "species") %>%  # <1>
  # Move 'family' column to the left of all other columns
  dplyr::relocate(family, .before = dplyr::everything())

# Look at the result of that
fish_df
```

1.  The 'by' argument accepts a vector of column names found in both data tables
