## Exploring the Data

:::{.callout-tip icon="false" collapse="false"}
### {{< fa brain >}} Learning Objectives

After completing this topic you will be able to: 

- <u>Extract</u> typical data structure information

:::

Before beginning any code operations, it's important to get a sense for the data. Characteristics like the dimensions of the dataset, the column names, and the type of information stored in each column are all crucial pre-requisites to knowing what tools can or should be used on the data.

Checking the data structure is one way of getting a lot of this high-level information.

```{r explore-libs}
#| include: false

library(lterdatasampler)
```

```{r structure}
# Check dataset structure
str(pie_crab)
```

For data that are primarily numeric, you may find data summary functions to be valuable. Note that most functions of this type do not provide useful information on text columns so you'll need to find that information elsewhere.

```{r summary}
# Get a simple summary of the data
summary(pie_crab)
```

For text columns it can sometimes be useful to simply look at the unique entries in a given column and sort them alphabetically for ease of parsing.

```{r sort-unique}
# Look at the sites included in the data
sort(unique(pie_crab$site))
```

For those of you who think more visually, a histogram can be a nice way of examining numeric data. There are simple histogram functions in the 'base' packages of most programming languages but it can sometimes be worth it to use those from special libraries because they can often convey additional detail.

```{r multi-hist}
#| message: false
#| fig-align: center
#| fig-width: 4
#| fig-height: 3

# Load the psych library
library(psych)

# Get the histogram of crab "size" (carapace width in mm)
psych::multi.hist(pie_crab$size)
```

::: {.callout-warning icon="false"}
#### Discussion: Data Exploration Tools

With a group of 4-5 others discuss the following questions:

-   What tools do you use when exploring new data?
-   Do you already use any of these tools to explore your data?
    -   If you do, why do you use them?
    -   If not, where do you think they might be valuable to include?
-   What value--if any--do you see in including these exploratory efforts in your code workflow?

:::
