## Types of Spatial Data

:::{.callout-tip icon="false" collapse="false"}
### {{< fa brain >}} Learning Objectives

After completing this topic you will be able to: 

- <u>Define</u> the difference between the two major types of spatial data
- <u>Manipulate</u> spatial data with R

:::

```{r}
#| include: false

# !!! NOTE TO SSECR DESIGN TEAM !!!

# This module uses raster and vector data that *are not widely available*
# You _must_ run run the 'prep-data_spatial.R' script in the "scripts/" directory before you will be able to render this module
# Reach out to Nick Lyon if questions exist
```

There are two main types of spatial data: vector and raster. Both types (and the packages they require) are described in the tabs below.

:::{.panel-tabset}
### Vector Data

Vector data are stored as polygons. Essentially vector data are a set of points and--sometimes--the lines between them that define the edges of a shape. They may store additional data that is retained in a semi-tabular format that relates to the polygon(s) but isn't directly stored in them.

Common vector data types include shape files or GeoJSONs.

```{r read-vector}
#| message: false
#| output: false

# Load needed library
library(sf)

# Read in shapefile
nc_poly <- sf::st_read(dsn = file.path("data", "nc_borders.shp")) #<1>
```
1. Note that even though we're only specifying the ".shp" file in this function you _must_ also have the associated files in that same folder. In this case that includes a ".dbf", ".prj", and ".shx", though in other contexts you may have others.

Once you have read in the shapefile, you can check its structure as you would any other data object. Note that the object has both the 'data.frame' class and the 'sf' ("simple features") class. In this case, the shapefile relates to counties in North Carolina and some associated demographic data in those counties.

```{r str-vector}
# Check structure
str(nc_poly)
```

If desired, we could make a simple R base plot-style map. In this case we'll do it based on just the county areas so that the polygons are filled with a color corresponding to how large the county is.

```{r plot-vector}
#| fig-align: center

# Make a graph
plot(nc_poly["AREA"], axes = T)
```

### Raster Data

Raster data are stored as values in pixels. The resolution (i.e., size of the pixels) may differ among rasters but in all cases the data are stored at a per-pixel level.

Common raster data types include GeoTIFFs (.tif) and NetCDF (.nc) files.

```{r read-raster}
#| message: false

# Load needed library
library(terra)

# Read in raster
nc_pixel <- terra::rast(x = file.path("data", "nc_elevation.tif"))
```

Once you've read in the raster file you can check it's structure as you would any other object but the resulting output is much less informative than for other object classes.

```{r str-raster}
# Check structure
str(nc_pixel)
```

Regardless, now that we have the raster loaded we can make a simple graph to check out what sort of data is stored in it. In this case, each pixel is 3 arcseconds on each side (~0.0002° latitude/longitude) and contains the elevation (in meters) of that pixel.

```{r plot-raster}
#| fig-align: center

# Make a graph
terra::plot(nc_pixel)
```

:::