## Visualization for Exploration

:::{.callout-tip icon="false" collapse="false"}
### {{< fa brain >}} Learning Objectives

After completing this topic you will be able to: 

- <u>Explain</u> how data visualization can be used to explore data

:::

Exploratory data visualization is an important part of any scientific project. Before launching into analysis it is valuable to make some simple plots to scan the contents. These plots may reveal any number of issues, such as typos, sensor calibration problems or differences in the protocol over time.

These "fitness for use" visualizations are even more critical for synthesis projects. In synthesis, we are often re-purposing publicly available datasets to answer questions that differ from the original motivations for data collection. As a result, the metadata included with a published dataset may be insufficient to assess whether the data are useful for your group's question. Datasets may not have been carefully quality-controlled prior to publication and could include any number of 'warts' that can complicate analyses or bias results. Some of these idiosyncrasies you may be able to anticipate in advance (e.g. spelling errors in taxonomy) and we encourage you to explicitly test for those and rectify them during the data harmonization process. Others may come as a surprise.

During the early stages of a synthesis project, you will want to gain skill to quickly scan through large volumes of data. The figures you make will typically be for internal use only, and therefore have low emphasis on aesthetics.

### Exploratory Visualization Applications

Specific applications of exploratory data visualization include identifying:

:::{.panel-tabset}
### Dataset Coverage

This may include temporal, spatial, or taxonomic coverage, to name a few.

For example, the metadata might indicate a dataset covers the period 2010-2020. That could mean one data point in 2010 and one in 2020! This may not be useful for a time-series analysis.

### Metadata Errors

Do the units "make sense" with the figure? Typos in metadata do occur, so if you find yourself with elephants weighing only a few grams, it may be necessary to reach out to the dataset contact.

### Methodological Differences

Consider the following:

- Do the data from sequential years, replicate sites, different providers generally fall into the same ranges or is there sensor drift or changes in protocols that need to be addressed?
- A risk of synthesis projects is that you may find you are comparing apples to oranges across datasets, as the individual datasets included in your project were likely not collected in a coordinated fashion.
- A benefit of synthesis projects is you will typically have large volumes of data, collected from many locations or timepoints. This data volume can be leveraged to give you a good idea of how your response variable looks at a 'typical' location as well as inform your gestalt sense of how much site-to-site, study-to-study, or year-to-year variability is expected. In our experience, where one particular dataset, or time period, strongly differs from the others, the most common root cause is differences in methodology that need to be addressed in the data harmonization process. 

### Other Issues

In addition to those other uses of exploratory visualization, you may also find:

- Harmonization issues
    - Are all your datasets measured in units that can be converted to the same units?
    - If not, can you envision metrics (relative abundance? Effect size?) that would make datasets intercomparable?
- Some entire datasets cannot be used
- Parts of some datasets cannot be used
- Additional quality control is needed (e.g. filtering large outliers)

:::

Identifying all of these pieces of information is an important precursor to the data harmonization stage, where you will process the datasets you have selected into an analysis-ready format. Visualization is a--relatively--easy way of doing these checks!

::::{.callout-warning icon="false"}
### {{<fa rocket >}} Activity: Data Sleuth

In this activity, you'll play the role of data detective. You will have many potential datasets to look through. It is important to do it correctly, but you likely won't need or want to develop boutique code to examine each dataset, especially since some may be discarded after an initial pass.

As a project team, discuss the following points:

1. Decide on a structure for tracking results of exploratory data checks
    - Git issues? Additional columns in your team-data-inventory google sheet? Something else?
    - Draft a list of 'generic checks' you would want to apply to each dataset before inclusion in your synthesis
2. Use the `summarytools` and/or `datacleanr` packages to explore one exemplar dataset that you intend to include in your project
    - Discuss any issues you discover
    - Create a "to do" list for the exemplar dataset that details additional steps needed to make that dataset analysis ready (e.g. remove 1993 due to incomplete sampling, convert concentrations from mmols to mg/L, contact dataset providers to ask about anomalous values in April 2021)
    - Revise the list of 'generic checks' for remaining datasets as necessary
3. If you choose to save any images and/or code you used in your exploratory data visualization, decide on a naming convention and storage location
    - Will you add these files to your `.gitignore` or do you plan on committing them?
4. What additional plots would you ideally make that are not available through these generic tools?

:::{.panel-tabset}
### `summarytools` Package

```{r data-sleuth_summarytools}
#| eval: false

# Load the library
library(summarytools)

# Load data
dataset_1 <- read_csv("your_file_name_here.csv")

# View the data in your Rstudio environment
summarytools::view(summarytools::dfSummary(dataset_1), footnote = NA) # <1>

# Alternatively,save the results for viewing later, or to share with your team
print(summarytools::dfSummary(dataset_1), footnote = NA,
      file = 'dataset_01_summary.html')
```
1. Careful! Use lowercase 'v' in the `view` function of the `summarytools` package

### `datacleanr` Package

```{r data-sleuth_datacleanr}
#| eval: false

# Load the library
library(datacleanr)

# Load data
dataset_1 <- read_csv("your_file_name_here.csv")

# Launch the shiny app and view the data interactively
datacleanr::dcr_app(dataset_1)
```

:::

Both of these packages have extensive vignettes and online instructional materials. See [here](https://cran.r-project.org/web/packages/summarytools/vignettes/introduction.html) for one from `summarytools` and [here](https://the-hull.github.io/datacleanr/) for one from `datacleanr`.

::::
