## Graphing with `ggplot2`

You may already be familiar with the `ggplot2` package in R but if you are not, it is a popular graphing library based on [The Grammar of Graphics](https://bookshop.org/p/books/the-grammar-of-graphics-leland-wilkinson/1518348?ean=9780387245447). Every ggplot is composed of four elements:

1. A 'core' `ggplot` function call
2. Aesthetics
3. Geometries
4. Theme

Note that the theme component may be implicit in some graphs because there is a suite of default theme elements that applies unless otherwise specified. 

This module will use example data to demonstrate these tools but as we work through these topics you should <u>feel free to substitute a dataset of your choosing</u>! If you don't have one in mind, you can use the example dataset shown in the code chunks throughout this module. This dataset comes from the [`lterdatasampler` R package](https://lter.github.io/lterdatasampler/) and the data are about fiddler crabs (_Minuca pugnax_) at the [Plum Island Ecosystems (PIE) LTER](https://pie-lter.mbl.edu/) site.

```{r data-prep}
#| eval: false

#| message: false
#| warning: false

# Load needed libraries
library(tidyverse); library(lterdatasampler)

# Load the fiddler crab dataset
data(pie_crab)

# Check its structure
str(pie_crab)
```

With this dataset in hand, let's make a series of increasingly customized graphs to demonstrate some of the tools in `ggplot2`.

::::{.panel-tabset}
### 1. Starter Graph

Let's begin with a scatterplot of crab size on the Y-axis with latitude on the X. We'll forgo doing anything to the theme elements at this point to focus on the other three elements.

```{r gg-1}
#| eval: false


#| fig-align: center
#| fig-width: 9
#| fig-height: 4

ggplot(data = pie_crab, mapping = aes(x = latitude, y = size, fill = site)) + # <1>
  geom_point(pch = 21, size = 2, alpha = 0.5) # <2>
```
1. We're defining both the data and the X/Y aesthetics in this top-level bit of the plot. Also, note that each line ends with a plus sign
2. Because we defined the data and aesthetics in the `ggplot()` function call above, this geometry can assume those mappings without re-specifying

### 2. Custom Theme

We can improve on this graph by tweaking theme elements to make it use fewer of the default settings.

```{r gg-2}
#| eval: false


#| fig-align: center
#| fig-width: 9
#| fig-height: 4

ggplot(data = pie_crab, mapping = aes(x = latitude, y = size, fill = site)) +
  geom_point(pch = 21, size = 2, alpha = 0.5) +
  theme(legend.title = element_blank(), # <1>
        panel.background = element_blank(),
        axis.line = element_line(color = "black"))
```
1. All theme elements require these `element_...` helper functions. `element_blank` removes theme elements but otherwise you'll need to use the helper function that corresponds to the type of theme element (e.g., `element_text` for theme elements affecting graph text)

### 3. Multiple Geometries

We can further modify `ggplot2` graphs by adding _multiple_ geometries if you find it valuable to do so. Note however that <u>geometry order matters</u>! Geometries added later will be "in front of" those added earlier. Also, adding too much data to a plot will begin to make it difficult for others to understand the central take-away of the graph so you may want to be careful about the level of information density in each graph. Let's add boxplots behind the points to characterize the distribution of points more quantitatively.

```{r gg-3}
#| eval: false


#| fig-align: center
#| fig-width: 9
#| fig-height: 4

ggplot(data = pie_crab, mapping = aes(x = latitude, y = size, fill = site)) +
  geom_boxplot(pch = 21) + # <1>
  geom_point(pch = 21, size = 2, alpha = 0.5) +
  theme(legend.title = element_blank(), 
        panel.background = element_blank(),
        axis.line = element_line(color = "black"))
```
1. By putting the boxplot geometry first we ensure that it doesn't cover up the points that overlap with the 'box' part of each boxplot

### 4. Multiple Datasets

`ggplot2` also supports adding more than one data object to the same graph! While this module doesn't cover map creation, maps are a common example of a graph with more than one data object. Another common use would be to include both the full dataset and some summarized facet of it in the same plot.

Let's calculate some summary statistics of crab size to include that in our plot.

```{r gg-4-prep}
#| eval: false


#| message: false

# Load the supportR library
library(supportR)

# Summarize crab size within latitude groups
crab_summary <- supportR::summary_table(data = pie_crab, groups = c("site", "latitude"),
                                        response = "size", drop_na = TRUE)

# Check the structure
str(crab_summary)
```

With this data object in-hand, we can make a graph that includes both this and the original, un-summarized crab data. To better focus on the 'multiple data objects' bit of this example we'll pare down on the actual graph code.

```{r gg-4}
#| eval: false


#| fig-align: center
#| fig-width: 9
#| fig-height: 4

ggplot() + # <1>
  geom_point(pie_crab, mapping = aes(x = latitude, y = size, fill = site),
             pch = 21, size = 2, alpha = 0.2) + 
  geom_errorbar(crab_summary, mapping = aes(x = latitude, # <2>
                                            ymax = mean + std_error,
                                            ymin = mean - std_error),
                width = 0.2) +
  geom_point(crab_summary, mapping = aes(x = latitude, y = mean, fill = site),
             pch = 23, size = 3) + 
  theme(legend.title = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(color = "black"))
```
1. If you want multiple data objects in the same `ggplot2` graph you need to leave this top level `ggplot()` call _empty!_ Otherwise you'll get weird errors with aesthetics later in the graph
2. This geometry adds the error bars and it's important that we add it before the summarized data points themselves if we want the error bars to be 'behind' their respective points

::::
