## Multi-Panel Graphs

It is sometimes the case that you want to make a single graph file that has multiple panels. For many of us, we might default to creating the separate graphs that we want, exporting them, and then using software like Microsoft PowerPoint to stitch those panels into the single image we had in mind from the start. However, as all of us who have used this method know, this is hugely cumbersome when your advisor/committee/reviewers ask for edits and you now have to redo all of the manual work behind your multi-panel graph. 

Fortunately, there are two nice entirely scripted alternatives that you might consider: **Faceted graphs** and **Plot grids**. See below for more information on both.

:::{.panel-tabset}
### Facets

In a faceted graph, <u>every panel of the graph has the same aesthetics</u>. These are often used when you want to show the relationship between two (or more) variables but separated by some other variable. In synthesis work, you might show the relationship between your core response and explanatory variables but facet by the original study. This would leave you with one panel per study where each would show the relationship only at that particular study.

Let's check out an example.

```{r facet-1}
#| fig-align: center
#| fig-width: 6
#| fig-height: 6

ggplot(pie_crab, aes(x = date, y = size, color = site))+
  geom_point(size = 2) +
  facet_wrap(. ~ site) + # <1>
  theme_bw() +
  theme(legend.position = "none") # <2>
```
1. This is a `ggplot2` function that assumes you want panels laid out in a regular grid. There are other `facet_...` alternatives that let you specify row versus column arrangement. You could also facet by multiple variables by putting something to the left of the tilde
2. We can remove the legend because the site names are in the facet titles in the gray boxes

### Plot Grids

In a plot grid, <u>each panel is completely independent of all others</u>. These are often used in publications where you want to highlight several _different_ relationships that have some thematic connection. In synthesis work, your hypotheses may be more complicated than in primary research and such a plot grid would then be necessary to put all visual evidence for a hypothesis in the same location. On a practical note, plot grids are also a common way of circumventing figure number limits enforced by journals.

Let's check out an example that relies on the `cowplot` library.

```{r grid-1}
#| message: false
#| fig-align: center
#| fig-width: 9
#| fig-height: 5

# Load a needed library
library(cowplot)

# Create the first graph
crab_p1 <- ggplot(pie_crab, aes(x = site, y = size, fill = site)) + # <1>
  geom_violin() +
  coord_flip() + # <2>
  theme_bw() +
  theme(legend.position = "none")

# Create the second
crab_p2 <- ggplot(pie_crab, aes(x = air_temp, y = water_temp)) +
  geom_errorbar(aes(ymax = water_temp + water_temp_sd, ymin = water_temp - water_temp_sd),
                 width = 0.1) +
  geom_errorbarh(aes(xmax = air_temp + air_temp_sd, xmin = air_temp - air_temp_sd), # <3>
                 width = 0.1) +
  geom_point(aes(fill = site), pch = 23, size = 3) +
  theme_bw()

# Assemble into a plot grid
cowplot::plot_grid(crab_p1, crab_p2, labels = "AUTO", nrow = 1) # <4>
```
1. Note that we're assigning these graphs to objects!
2. This is a handy function for flipping X and Y axes without re-mapping the aesthetics
3. This geometry is responsible for _horizontal_ error bars (note the "h" at the end of the function name)
4. The `labels = "AUTO"` argument means that each panel of the plot grid gets the next sequential capital letter. You could also substitute that for a vector with labels of your choosing
:::
